From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-07-12 mechanics/paper.tex
Final changes

Reynolds' theory of relational parametricity \cite{reynolds82} tells
us that parametrically polymorphic programs automatically satisfy
invariance properties. Such invariance properties are often called
\emph{Free Theorems}, after Wadler \cite{wadler89theorems}, since they
follow ``for free'' from the types of programs, rather than through
detailed study of the program text itself. An illustrative example is
the free theorem for programs $f$ with the following type:
\begin{displaymath}
  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{Nat}
\end{displaymath}
Such programs takes lists of $\alpha$s, for any type $\alpha$, to
natural numbers. Using Reynolds' theory of relational parametricity,
Wadler showed that \emph{any} $f$ with this type satisfies the
following property:
\begin{displaymath}
  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~f~[\alpha]~l = f~[\beta]~(\mathrm{map}~g~l)
\end{displaymath}
Thus, any $f$ with the type given above is invariant under mapping
some arbitrary function $g$ over its input. Thinking in terms of
abstract data types and change of data representation, this free
theorem states that $f$ is invariant under change of data
representation from an arbitrary type $\alpha$ to another arbitrary
type $\beta$, via $g$.

The use of invariance under change to derive useful consequences is a
technique much older than programming languages. In physics,
\emph{Noether's theorem} \cite{noether} provides a general way to
derive conservation laws for physical systems from invariance
properties. Informally speaking, Noether's theorem states that every
continuous symmetry of a physical system implies the existence of a
corresponding conserved quantity. A conserved quantity is a function
of the state of the system that is constant for all time. We
illustrate the use of Noether's theorem with a simple example of a
classical system consisting of two particles of equal mass $m$
connected by a spring with spring constant $k$, such that the
particles are constrained to move one-dimensionally.

@@

Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on older
work by Kennedy \cite{kennedy94para}, have already presented a
polymorphic type system for expressing geometric invariance properties
similar to the translation invariance property in Equation
(\ref{eq:coupled-spring-translation-invariance}). In the present work,
we extend their type system by embedding it within System
F$\omega$. The benefits of this embedding are twofold. Firstly, by
using the richer type-level structure of System F$\omega$, we can
easily add useful indexed types like length indexed vectors and smooth
function types, both of which would have required special treatment in
the special purpose type system presented by Atkey \emph{et
  al.}. Secondly, by using System F$\omega$ as our base, we are able
to reuse the reflexive graph-based relationally parametric semantics
presented by Atkey \cite{atkey12relational}, extending it as
appropriate for our geometric setting.

@@

Noether's theorem applies to systems described using the formalism of
Lagrangian mechanics, a reformulation of Newtonian classical
mechanics. In this section we can only offer a very brief introduction
to the concepts of Lagrangian mechanics. The reader is referred to
other sources, such as Landau and Lifschitz \cite{landau60mechanics}
or Arnol'd \cite{arnoldXXmathematical} for the necessary background in
the mathematical theory of classical mechanics. Lagrangian mechanics,
and Noether's theorem, make crucial use of the Calculus of
Variations. Gelfand and Fomin \cite{gelfandXXcalculus} provide a good
introduction to this topic.

@@

Lagrangian mechanics is a branch of analytical mechanics, along with
Hamiltonian mechanics
\cite{landau60mechanics,arnoldXXmathematical}. Analytical mechanics
seeks to discover the underlying structure of Newtonian classical
mechanics by reformulating it in different terms.

@@

The \emph{principle of stationary action} states that the physically
realisable paths $q$ are the paths that yield stationary points (i.e.,
either minima or maxima) of the action $\mathcal{S}[q;a;b]$, for all
$a$ and $b$. Defining what is exactly meant by a stationary point of
the action is beyond the scope of this paper, and can be found in
standard references on the calculus of variations, such as Gelfand and
Fomin \cite{gelfandXXcalculus}. The key point for our purposes is that
a path $q$ satisfies the principle of stationary action if and only if
it satisfies a system of ODEs derived from the Lagrangian $L$, called
the Euler-Lagrange equations:
\begin{equation}\label{eq:euler-lagrange}
  \frac{d}{dt}\frac{\partial L}{\partial \dot{q_i}} - \frac{\partial L}{\partial q_i} = 0
\end{equation}
The ODEs derived from the Lagrangian for a classical mechanical system
are called the \emph{equations of motion} for the system.

The notation in the Euler-Lagrange equation, while standard, is not
exactly clear, especially to a programming language theorist used to a
more careful treatment of variables. The two partial derivatives
$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
hence the appearance of the second derivatives $\ddot{q_i}$ in the
equations of motion. Sussman and Wisdom \cite{sussman} criticise the
standard presentation of Lagrangian mechanics for this kind of
confusing notation, and define a clearer notation based on the Scheme
programming language. We elect to stick with the standard notation as
used in physics textbooks just for familiarity.

@@

\begin{proof}
  E.g., Gelfand and Fomin \cite{gelfandXXcalculus}, Section 20.
\end{proof}

\begin{example}
  In Example~\ref{ex:invariance-examples}, we noted that the
  Lagrangians (\ref{eq:coupled-spring-lagrangian}) and
  (\ref{eq:simple-g-lagrangian}) are both invariant under arbitrary
  translations in time. In terms of continuous invariance, their
  corresponding actions are invariant under the families
  $\Phi_\epsilon(t) = t + \epsilon$ and $\Psi_\epsilon(q) =
  q$. Plugging these definitions into
  Equation~(\ref{eq:conservation-law}), we learn that, for these
  systems:
  \begin{displaymath}
    \frac{d}{dt}~\left(L - \sum_{i=1}^2 \dot{q_i}\frac{\partial L}{\partial \dot{q_i}} \right) = 0
  \end{displaymath}
  For the spring-coupled particles Lagrangian,
  (\ref{eq:coupled-spring-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2)^2 \right) = 0
  \end{displaymath}
  and for the single particle under gravity Lagrangian
  (\ref{eq:simple-g-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x}^2 + \dot{y}^2) + mgy\right) = 0
  \end{displaymath}
  In both cases, we have discovered that, as a consequence of
  invariance under translation in time, the total energy of the system
  is conserved (note that the $-$ signs in the original Lagrangians
  have turned into $+$s). In general, invariance under translation in
  time implies, via Noether's theorem, conservation of energy.
\end{example}

@@

We derive invariance properties from types by constructing a
relationally parametric model of System F$\omega$, which we then
extend with suitable kind-, type-, and term-level constants for
writing invariant Lagrangians. Our relationally parametric model is
based on the reflexive graph model presented by
Atkey~\cite{atkey12relational} (based on previous work on reflexive
graph models of relational parametricity by Robinson and
Rosolini~\cite{robinson94} and Hasegawa~\cite{hasegawa94}). In
reflexive graph models of relational parametricity, the kinds of
System F$\omega$ are interpreted as reflexive graphs, while types are
interpreted as morphisms of reflexive graphs. As demonstrated by
Atkey, uniformly interpreting kinds as reflexive graphs allows for a
straightforward interpretation of higher kinds like $* \to *$. In the
present setting, reflexive graphs will allow us to accomodate
arbitrary groupoids as kinds
(Section~\ref{sec:groupoid-kinds}). Groupoid kinds are the key
technical tool that we need to derive the kinds of geometric
invariance properties we need for Noether's theorem.

@@

We deviate from the standard presentation of System F$\omega$ in that
we distinguish between large and small kinds, and only allow
quantification over small kinds. In particular, the kind of types $*$
is \emph{not} small. The reason for this is that we desire a simple
set-theoretic model. However, requiring that we have a set of sets
closed under large products, in order to interpret quantification over
the kind $*$, would force us to use an intuitionistic metatheory
\cite{pitts}. The basic results of the calculus of variations that we
are relying on, such as the derivation of the Euler-Lagrange
equations, use excluded middle in their proofs, so an intuitionistic
metatheory is not an option. We therefore just assume that we have a
set of ``small'' sets, which includes the real numbers and is closed
under subsets, products, function space and comprehension.

@@

The well-kinded types, type equalities, and well-typed terms of System
F$\omega$ are shown in Figures~\ref{fig:types},
\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Note that
typing contexts $\Gamma$ only contain well-kinded types of kind
$*$. Terms also have an equational theory, with $\beta\eta$-laws for
functions, products, and universal quantification, which we omit. Our
presentation of System F$\omega$ is entirely standard (see, e.g.,
Pierce~\cite{pierce01types}), except for the restriction to small
kinds in the formation of universal kinds $\forall
\alpha\mathord:\kappa. A$. As is the case for kinds, we will extend
the types, type equalities and terms in
Sections~\ref{sec:discrete-kinds},~\ref{sec:groupoid-kinds}
and~\ref{sec:types-for-classical-mech}, below.

@@

It will be useful for our examples to have the following trigonometric
functions, exponential function and division operation. The types of
each of these functions describes some of their invariance properties,
for example the periodicity of the $\sin$ and $\cos$ functions. A
technical problem with our current system is that we do not account
for the non-definedness of division at $0$. However, division will be
useful in many of our examples, so we take a pragmatic approach, in
common with most physics textbooks, and informally treat division as
total function.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle) \\
    (/) &:& \forall g_1, g_2 \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}\langle g_2, 0\rangle, \mathbb{R}\langle g_1g_2^{-1}, 0 \rangle)
  \end{array}
\end{displaymath}

Finally, we include two primitive combinators for dealing with $n$-ary
products of vectors. Both of these combinators will be useful when
defining Lagrangians that are generic in the number of particles.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{sum} &:& \forall n \mathord: \mathsf{Nat}, g \mathord: \mathsf{GL}(1). C^\infty(\mathit{vec}~n~(\mathbb{R}\langle g, 0 \rangle), \mathbb{R}\langle g, 0\rangle) \\
    \mathrm{map} &:& \forall n \mathord: \mathsf{Nat}. C^\infty(Z \times X,Y) \to C^\infty(Z \times \mathrm{vec}~n~X, \mathrm{vec}~n~Y) \\
    \mathrm{cross} &:& 
    \begin{array}[t]{@{}l}
      \forall m,n \mathord: \mathsf{Nat}.\\
      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~(X \times Y))
    \end{array}
  \end{array}
\end{displaymath}
The combinator $\mathrm{sum}$ sums the list of real numbers it is
given, while $\mathrm{cross}$ takes the cartesian product of two
vectors of elements of cartesian spaces, and $\mathrm{map}$ maps the
given smooth function over a vector. The multiplication of natural
numbers $m*n$ is straightforwardly defined in terms of the
$\mathit{natrec}$ recursion operator. We will use $\mathrm{cross}$ in
Section~\ref{sec:examples} to define Lagrangians in terms of the
interactions between systems of particles.

@@

  \medskip
  \textbf{Division, sum, and cross}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}\langle G_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \frac{e_1}{e_2} : \mathbb{R}\langle G_1G_2^{-1}, 0 \rangle}

@@

\begin{figure}
  \centering
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \elab{z}_\Delta & = & \mathrm{const}~z \\
      \elab{x}_\Delta & = & \underline{\mathrm{project}_\Delta}(x) \\
      \elab{\mathrm{let}~\vec{x} = e_1\textrm{ in }e_2}_\Delta & = & \mathrm{pair}~\mathrm{id}~\elab{e_1}_\Delta \comp \elab{e_2}_{\Delta, \vec{x}} \\
      \elab{(e_1,...,e_n)}_\Delta &=& \underline{\mathrm{pair}_n}(\elab{e_1}_\Delta, ..., \elab{e_n}_\Delta) \\
      \elab{e_1(e_2)}_\Delta &=& \elab{e_2}_\Delta \comp e_1 \\
      \elab{\vec{c}}_\Delta &=& \mathrm{const}~\vec{c} \\
      \elab{0}_\Delta &=& \mathrm{const}~0 \\
      \elab{e_1 + e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (+) \\
      \elab{e_1 - e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (-) \\
      \elab{e_1 e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (*) \\
      \elab{e_1 \cdot e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (\cdot) \\
      \elab{\sin~e}_\Delta &=& \elab{e}_\Delta \comp \sin \\
      \elab{\cos~e}_\Delta &=& \elab{e}_\Delta \comp \cos \\
      \elab{\mathrm{exp}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{exp} \\
      \elab{e_1 / e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (/) \\
      \elab{\mathrm{sum}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{sum} \\
      \elab{\mathrm{map}~(x.~e_1)~e_2}_\Delta &=& \mathrm{pair}~\mathrm{id}~\elab{e_2}_\Delta \comp (\mathrm{map}~\elab{e_1}_{\Delta,x}) \\
      \elab{\mathrm{cross}~e_1~e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp \mathrm{cross}
    \end{array}
  \end{displaymath}
  \caption{Desugaring of the surface syntax}
  \label{fig:desugar}
\end{figure}

@@

>>>>

Reynolds' theory of relational parametricity \cite{reynolds83types}
tells us that parametrically polymorphic programs automatically
satisfy invariance properties. Such invariance properties are often
called \emph{Free Theorems}, after Wadler \cite{wadler89theorems},
since they follow ``for free'' from the types of programs, rather than
through detailed study of the program text itself. An illustrative
example is the free theorem for programs $f$ with the following type:
\begin{displaymath}
  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{Nat}
\end{displaymath}
Such programs takes lists of $\alpha$s, for any type $\alpha$, to
natural numbers. Using Reynolds' theory of relational parametricity,
Wadler showed that \emph{any} $f$ with this type satisfies the
following property:
\begin{displaymath}
  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~f~[\alpha]~l = f~[\beta]~(\mathrm{map}~g~l)
\end{displaymath}
Thus, any $f$ with the type given above is invariant under mapping
some arbitrary function $g$ over its input. Thinking in terms of
abstract data types and change of data representation, this free
theorem states that $f$ is invariant under change of data
representation from an arbitrary type $\alpha$ to another arbitrary
type $\beta$, via $g$.

The use of invariance under change to derive useful consequences is a
technique much older than programming languages. In physics,
\emph{Noether's theorem} \cite{noether71} provides a general way to
derive conservation laws for physical systems from invariance
properties. Informally speaking, Noether's theorem states that every
continuous symmetry of a physical system implies the existence of a
corresponding conserved quantity. A conserved quantity is a function
of the state of the system that is constant for all time. We
illustrate the use of Noether's theorem with a simple example of a
classical system consisting of two particles of equal mass $m$
connected by a spring with spring constant $k$, such that the
particles are constrained to move one-dimensionally.

@@

Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on older
work by Kennedy \cite{kennedy97relational}, have already presented a
polymorphic type system for expressing geometric invariance properties
similar to the translation invariance property in Equation
(\ref{eq:coupled-spring-translation-invariance}). In the present work,
we extend their type system by embedding it within System
F$\omega$. The benefits of this embedding are twofold. Firstly, by
using the richer type-level structure of System F$\omega$, we can
easily add useful indexed types like length indexed vectors and smooth
function types, both of which would have required special treatment in
the special purpose type system presented by Atkey \emph{et
  al.}. Secondly, by using System F$\omega$ as our base, we are able
to reuse the reflexive graph-based relationally parametric semantics
presented by Atkey \cite{atkey12relational}, extending it as
appropriate for our geometric setting.

@@

Noether's theorem applies to systems described using the formalism of
Lagrangian mechanics, a reformulation of Newtonian classical
mechanics. In this section we can only offer a very brief introduction
to the concepts of Lagrangian mechanics. The reader is referred to
other sources, such as Landau and Lifschitz \cite{landau60mechanics}
or Arnol'd \cite{arnold89mathematical} for the necessary background in
the mathematical theory of classical mechanics. Lagrangian mechanics,
and Noether's theorem, make crucial use of the Calculus of
Variations. Gelfand and Fomin \cite{gelfand00calculus} provide a good
introduction to this topic.

@@

Lagrangian mechanics is a branch of analytical mechanics, along with
Hamiltonian mechanics
\cite{landau60mechanics,arnold89mathematical}. Analytical mechanics
seeks to discover the underlying structure of Newtonian classical
mechanics by reformulating it in different terms.

@@

The \emph{principle of stationary action} states that the physically
realisable paths $q$ are the paths that yield stationary points (i.e.,
either minima or maxima) of the action $\mathcal{S}[q;a;b]$, for all
$a$ and $b$. Defining what is exactly meant by a stationary point of
the action is beyond the scope of this paper, and can be found in
standard references on the calculus of variations, such as Gelfand and
Fomin \cite{gelfand00calculus}. The key point for our purposes is that
a path $q$ satisfies the principle of stationary action if and only if
it satisfies a system of ODEs derived from the Lagrangian $L$, called
the Euler-Lagrange equations:
\begin{equation}\label{eq:euler-lagrange}
  \frac{d}{dt}\frac{\partial L}{\partial \dot{q_i}} - \frac{\partial L}{\partial q_i} = 0
\end{equation}
The ODEs derived from the Lagrangian for a classical mechanical system
are called the \emph{equations of motion} for the system.

The notation in the Euler-Lagrange equation, while standard, is not
exactly clear, especially to a programming language theorist used to a
more careful treatment of variables. The two partial derivatives
$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
hence the appearance of the second derivatives $\ddot{q_i}$ in the
equations of motion. Sussman and Wisdom \cite{sussman01structure}
criticise the standard presentation of Lagrangian mechanics for this
kind of confusing notation, and define a clearer notation based on the
Scheme programming language. We elect to stick with the standard
notation as used in physics textbooks just for familiarity.

@@

\begin{proof}
  E.g., Gelfand and Fomin \cite{gelfand00calculus}, Section 20.
\end{proof}

\begin{example}\label{ex:conservation-laws}
  In Example~\ref{ex:invariance-examples}, we noted that the
  Lagrangians (\ref{eq:coupled-spring-lagrangian}) and
  (\ref{eq:simple-g-lagrangian}) are both invariant under arbitrary
  translations in time. In terms of continuous invariance, their
  corresponding actions are invariant under the families
  $\Phi_\epsilon(t) = t + \epsilon$ and $\Psi_\epsilon(q) =
  q$. Plugging these definitions into
  Equation~(\ref{eq:conservation-law}), we learn that, for these
  systems:
  \begin{displaymath}
    \frac{d}{dt}~\left(L - \sum_{i=1}^2 \dot{q_i}\frac{\partial L}{\partial \dot{q_i}} \right) = 0
  \end{displaymath}
  For the spring-coupled particles Lagrangian,
  (\ref{eq:coupled-spring-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2)^2 \right) = 0
  \end{displaymath}
  and for the single particle under gravity Lagrangian
  (\ref{eq:simple-g-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x}^2 + \dot{y}^2) + mgy\right) = 0
  \end{displaymath}
  In both cases, we have discovered that, as a consequence of
  invariance under translation in time, the total energy of the system
  is conserved (note that the $-$ signs in the original Lagrangians
  have turned into $+$s). In general, invariance under translation in
  time implies, via Noether's theorem, conservation of energy.
\end{example}

@@

We derive invariance properties from types by constructing a
relationally parametric model of System F$\omega$, which we then
extend with suitable kind-, type-, and term-level constants for
writing invariant Lagrangians. Our relationally parametric model is
based on the reflexive graph model presented by
Atkey~\cite{atkey12relational} (based on previous work on reflexive
graph models of relational parametricity by Robinson and
Rosolini~\cite{robinson94reflexive} and
Hasegawa~\cite{hasegawa94relational}). In reflexive graph models of
relational parametricity, the kinds of System F$\omega$ are
interpreted as reflexive graphs, while types are interpreted as
morphisms of reflexive graphs. As demonstrated by Atkey, uniformly
interpreting kinds as reflexive graphs allows for a straightforward
interpretation of higher kinds like $* \to *$. In the present setting,
reflexive graphs will allow us to accomodate arbitrary groupoids as
kinds (Section~\ref{sec:groupoid-kinds}). Groupoid kinds are the key
technical tool that we need to derive the kinds of geometric
invariance properties we need for Noether's theorem.

@@

We deviate from the standard presentation of System F$\omega$ in that
we distinguish between large and small kinds, and only allow
quantification over small kinds. In particular, the kind of types $*$
is \emph{not} small. The reason for this is that we desire a simple
set-theoretic model. However, requiring that we have a set of sets
closed under large products, in order to interpret quantification over
the kind $*$, would force us to use an intuitionistic metatheory
\cite{pitts87polymorphism}. The basic results of the calculus of
variations that we are relying on, such as the derivation of the
Euler-Lagrange equations, use excluded middle in their proofs, so an
intuitionistic metatheory is not an option. We therefore just assume
that we have a set of ``small'' sets, which includes the real numbers
and is closed under subsets, products, function space and
comprehension.

@@

The well-kinded types, type equalities, and well-typed terms of System
F$\omega$ are shown in Figures~\ref{fig:types},
\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Note that
typing contexts $\Gamma$ only contain well-kinded types of kind
$*$. Terms also have an equational theory, with $\beta\eta$-laws for
functions, products, and universal quantification, which we omit. Our
presentation of System F$\omega$ is entirely standard (see, e.g.,
Pierce~\cite{pierce02types}), except for the restriction to small
kinds in the formation of universal kinds $\forall
\alpha\mathord:\kappa. A$. As is the case for kinds, we will extend
the types, type equalities and terms in
Sections~\ref{sec:discrete-kinds},~\ref{sec:groupoid-kinds}
and~\ref{sec:types-for-classical-mech}, below.

@@

It will be useful for our examples to have the following trigonometric
functions, exponential function and division operation. The types of
each of these functions describes some of their invariance properties,
for example the periodicity of the $\sin$ and $\cos$ functions. A
technical problem with our current system is that we do not account
for the non-definedness of division at $0$, nor do we account for the
square root of negative numbers. However, division and square root
will be useful in our examples in Section~\ref{sec:examples}, so we
take a pragmatic approach, in common with most physics textbooks, and
informally treat them as total functions.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle) \\
    (/) &:& \forall g_1, g_2 \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}\langle g_2, 0\rangle, \mathbb{R}\langle g_1g_2^{-1}, 0 \rangle) \\
    \mathrm{sqrt} &:& \forall g \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g^2, 0 \rangle, \mathbb{R}\langle g, 0 \rangle)
  \end{array}
\end{displaymath}

Finally, we include three primitive combinators for dealing with
$n$-ary products of vectors. These combinators will be useful when
defining Lagrangians that are generic in the number of particles.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{sum} &:& \forall n \mathord: \mathsf{Nat}, g \mathord: \mathsf{GL}(1). C^\infty(\mathit{vec}~n~(\mathbb{R}\langle g, 0 \rangle), \mathbb{R}\langle g, 0\rangle) \\
    \mathrm{map} &:& \forall n \mathord: \mathsf{Nat}. C^\infty(Z \times X,Y) \to C^\infty(Z \times \mathrm{vec}~n~X, \mathrm{vec}~n~Y) \\
    \mathrm{cross} &:& 
    \begin{array}[t]{@{}l}
      \forall m,n \mathord: \mathsf{Nat}.\\
      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~(X \times Y))
    \end{array}
  \end{array}
\end{displaymath}
The combinator $\mathrm{sum}$ sums the list of real numbers it is
given, while $\mathrm{cross}$ takes the cartesian product of two
vectors of elements of cartesian spaces, and $\mathrm{map}$ maps the
given smooth function over a vector. The multiplication of natural
numbers $m*n$ is straightforwardly defined in terms of the
$\mathit{natrec}$ recursion operator. We will use $\mathrm{cross}$ in
Section~\ref{sec:examples} to define Lagrangians in terms of the
interactions between systems of particles.

@@

  \medskip
  \textbf{Division, square root, sum, and cross}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}\langle G_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \frac{e_1}{e_2} : \mathbb{R}\langle G_1G_2^{-1}, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle G^2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sqrt}~e : \mathbb{R}\langle G, 0 \rangle}

@@

\begin{figure}
  \centering
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \elab{z}_\Delta & = & \mathrm{const}~z \\
      \elab{x}_\Delta & = & \underline{\mathrm{project}_\Delta}(x) \\
      \elab{\mathrm{let}~\vec{x} = e_1\textrm{ in }e_2}_\Delta & = & \mathrm{pair}~\mathrm{id}~\elab{e_1}_\Delta \comp \elab{e_2}_{\Delta, \vec{x}} \\
      \elab{(e_1,...,e_n)}_\Delta &=& \underline{\mathrm{pair}_n}(\elab{e_1}_\Delta, ..., \elab{e_n}_\Delta) \\
      \elab{e_1(e_2)}_\Delta &=& \elab{e_2}_\Delta \comp e_1 \\
      \elab{\vec{c}}_\Delta &=& \mathrm{const}~\vec{c} \\
      \elab{0}_\Delta &=& \mathrm{const}~0 \\
      \elab{e_1 + e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (+) \\
      \elab{e_1 - e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (-) \\
      \elab{e_1 e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (*) \\
      \elab{e_1 \cdot e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (\cdot) \\
      \elab{\sin~e}_\Delta &=& \elab{e}_\Delta \comp \sin \\
      \elab{\cos~e}_\Delta &=& \elab{e}_\Delta \comp \cos \\
      \elab{\mathrm{exp}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{exp} \\
      \elab{e_1 / e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (/) \\
      \elab{\mathrm{sqrt}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{sqrt} \\
      \elab{\mathrm{sum}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{sum} \\
      \elab{\mathrm{map}~(x.~e_1)~e_2}_\Delta &=& \mathrm{pair}~\mathrm{id}~\elab{e_2}_\Delta \comp (\mathrm{map}~\elab{e_1}_{\Delta,x}) \\
      \elab{\mathrm{cross}~e_1~e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp \mathrm{cross}
    \end{array}
  \end{displaymath}
  \caption{Desugaring of the surface syntax}
  \label{fig:desugar}
\end{figure}

@@

Our first example is of a simple Lagrangian for a single free particle
in $3$-dimensional space, with no external forces. We define the
Lagrangian using the syntax for smooth terms we defined in the
previous section, in the following kinding, typing and cartesian space
contexts:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}l}
    \Theta = & t_t : \mathsf{T}(1), t_x : \mathsf{T}(3), o : \mathsf{O}(3) \\
    \Gamma = & m : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle} \\
    \Delta = & t : \mathbb{R}\langle 1, t_t \rangle, x : \mathbb{R}^3\langle \mathit{ortho}_3(o), t_x \rangle, \dot{x} : \mathbb{R}^3\langle \mathit{ortho}_3(o), 0 \rangle
  \end{array}
\end{displaymath}
where $o$ represents an arbitrary orthogonal transformation of the
space, $t_t$ represents a translation in time, $t_x$ represents a
$3$-dimensional translation in space, $m$ is the constant mass of the
free particle, $t$ is the current time, $x$ is the current position of
the particle, and $\dot{x}$ is the current velocity.

We write the Lagrangian like so, which in the absence of external
forces just consists of the kinetic energy term:
\begin{displaymath}
  L = \frac{1}{2}m(\dot{x} \cdot \dot{x}) : \mathbb{R}\langle 1, 0 \rangle
\end{displaymath}

By the fundamental theorem of logical relations for our calculus
(Property~\ref{property:semantic-terms}), we can derive the following
free theorems, one each for the three group parameters $t_t$, $t_x$
and $o$. Quantifying over translations in time, $t_t$, gives us
invariance under translation in time:
\begin{displaymath}
  \forall t_t \in \mathbb{R}.~\sem{L}(t + t_t, \vec{x}, \dot{\vec{x}}) = \sem{L}(t, \vec{x}, \dot{\vec{x}})
\end{displaymath}
As we saw in Example~\ref{ex:conservation-laws}, Noether's theorem
tells us that invariance under translation in time yields conservation
of energy. Likewise, quantifying over translations in space, $t_x$,
gives us invariance under translation in space:
\begin{displaymath}
  \forall \vec{t_x} \in \mathbb{R}^3.~\sem{L}(t, \vec{x} + \vec{t_x}, \dot{\vec{x}}) = \sem{L}(t, \vec{x}, \dot{\vec{x}})
\end{displaymath}
Similar to the example of the spring-coupled particles in the
introduction, invariance under translations in space yields
conservation of linear momentum.
